### 2014-07-25 Paper_v1.tex
Updated text

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large An Effective Model of the Retinoic Acid induced HL-60 Differentiation Circuit}}
\vspace{0.05in}
{\par \centering \large{Ryan Tasseff$^{1}$, 
Holly A. Jensen$^{1}$, 
Johanna Congleton$^{2}$, Andrew Yen$^{2}$ and Jeffrey D. Varner$^{1,*}$}}
\vspace{0.10in}
{\par \centering \large{$^{1}$~School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \large{$^{2}$~Department of Biomedical Sciences}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Effective HL60 differentiation circuit}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Processes}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}

@@

\section*{Discussion}

@@

\subsection*{Formulation and solution of the model equations.}
We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell free
metabolic networks:
\begin{eqnarray}
	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of
enzymes in the model. The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ 
denotes the rate of reaction $j$. Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters 
$\mathbf{k}$ ($\mathcal{K}\times{1}$).
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. If 
$\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates
metabolite $i$ is not connected with reaction $j$. Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. The system material balances were subject to the
initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).

@@

The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
For each enzyme, we used a rule based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
If an enzyme was activated by $m$ metabolites, we modeled this activation as:
\begin{equation}
	v_{j} = \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
\end{equation}where $0\leq f_{ij}\left(x\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
\begin{equation}
	v_{j} = 1 - \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
\begin{equation}
	v_{j} = \min\left(u_{j},d_{j}\right)
\end{equation}where:
\begin{eqnarray}
	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right) \\
	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(x\right),\hdots,f_{nj}\left(x\right)\right)
\end{eqnarray}The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating, and inhibitory factors for enzyme $j$. 
If an enzyme had no allosteric factors, we set $v_{j} = 1$.
There are many possible functional forms for $0\leq f_{ij}\left(x\right)\leq 1$. 
However, in this study, each individual transfer function took the form:
\begin{equation}\label{eqn:control-factor}
	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}x_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}x_{j}^{\eta}}
\end{equation}where $x_{j}$ denotes the abundance of metabolite $j$, and $\kappa_{ij}$ and $\eta$ are control parameters. 
The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
The model equations were encoded using the Octave programming language, and solved using the LSODE routine in Octave \citep{Octave:2014}.

\subsection*{Estimation of model parameters from experimental data.}

\section*{Acknowledgements}
This study was supported by the National Science Foundation GK12 award (DGE-1045513) 
and by the National Science Foundation CAREER award (FILLMEIN).

>>>>

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large An Effective Model of the Core Retinoic Acid induced HL-60 Differentiation Circuit}}
\vspace{0.05in}
{\par \centering \large{Ryan Tasseff$^{1}$, 
Holly A. Jensen$^{1}$, 
Johanna Congleton$^{2}$, Andrew Yen$^{2}$ and Jeffrey D. Varner$^{1,*}$}}
\vspace{0.10in}
{\par \centering \large{$^{1}$~School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \large{$^{2}$~Department of Biomedical Sciences}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Effective HL60 differentiation circuit}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Processes}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
The complexity of human differentiation networks makes analysis of their architecture and function challenging.  
Analysis of model systems, such as the lineage-uncommitted human myeloblastic leukemia cell line HL-60, 
could inform the analysis of more complex programs. 
HL-60 cells undergo myeloid differentiation along with G1/G0-arrest when exposed to All-Trans Retinoic Acid (ATRA). 
One of the defining features of this program is slowly induced, persistent MAPK activation.
The molecular mechanisms of ATRA-induced commitment, arrest and functional differentiation are only partially understood.
In this study, we explored the ATRA-inducible c-Raf, also known as Raf1 (Raf), 
interactome to determine the functional and regulatory architecture responsible for persistent MAPK activation in HL-60 cells. 
To better understand the role of Raf in the ATRA-induced differentiation program, 
we constructed an ensemble of mechanistic, mathematical models of the transcriptional and post-translational events driving ATRA-induced MAPK activation. 
Bifurcation analysis of this model predicted bistability in ppERK levels as a function of ATRA forcing. 
A functional consequence of this was the ability to lock the MAPK cascade into a self-sustaining activated state, even after ATRA removal.
These simulations were then qualitatively validated with ATRA washout experiments. 
Taken together, this study provided further details on sustained MAPK activation, mechanistic insight into cellular memory,
and proof-of-concept that a
combination of experimental and computational methods is an effective strategy for dissecting complex intracellular signaling programs.

@@

\section*{Discussion}
In this study, we explored the functional and regulatory architecture responsible for ATRA-induced MAPK activation in HL-60 cells.
Retinoic acid is an embryonic morphogen, a cancer chemotherapeutic agent, as well as potent regulator of cell differentiation and proliferation. 
Using a human lineage-uncommitted myelo-monocytic precursor cell, 
it has been shown that ATRA-induced phenotypic conversion and G0 arrest depend on eliciting a hyperactive prolonged signal along the Raf/MEK/ERK axis \cite{Yen1998}. 
In particular, signaling involves two positive loops, which incorporates the BLR1 receptor and Raf kinase, putatively activated by S621 phosphorylation, 
that is seminal to sustaining the signal needed for differentiation \cite{Wang2008}. 
In genetically engineered BLR1 KO cells crippled for signaling and consequentially ability to differentiate, ectopic expression of Raf activity restores the cell's ability to differentiate. 
While the canonical paradigm of MAPK signaling is a transient signal that is mitogenic, a prolonged signal elicits cell differentiation and G0 arrest. 
Interest thus becomes focused on whether there is a Raf partner that is responsible for the prolonged signal in ATRA-treated cells. 
Surveying known MAPK regulatory molecules using Western blotting and IP we found that ATRA increased the expression of Vav1 as well as its immunoprecipitation with Raf phosphorylated at S621. 
A parsimonious effective mathematical model showed that a Raf-Siganlsome-mediated positive feedback loop accounted for sustained MAPK signaling and aspects of phenotypic conversion and G0 arrest. 
The significant finding was thus that ATRA-regulated Vav1 expression and partnering with Raf could drive sustained signaling and phenotypic shift. The model also predicted the observed hysteresis effect of ATRA where the cell retains a memory of transient ATRA exposure, betrayed by activated signaling and by accelerated differentiation upon re-exposure to ATRA.

@@

\subsection*{Formulation and solution of the model equations.}
We used ordinary differential equations (ODEs) to model the time evolution of the circuit components:
\begin{equation}
	\frac{dx_{i}}{dt}  = \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) - \left(\mu + k_{d,i}\right)x_{i}\qquad{i=1,2,\hdots,\mathcal{M}}\\
\end{equation}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of circuit components, $k_{d,i}$ denotes a first-order non-specific degradation constant
for species i, and $\mu$ denotes the specific growth rate of HL-60 cells.
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. If 
$\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates
metabolite $i$ is not connected with reaction $j$. 
The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$. 
Typically, reaction $j$ is a non-linear function of network state, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$).
The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.

@@

The control term $v_{j}$ depended upon the combination of factors which influenced the abundance or activity of $x_{i}$.
For each species, we used a rule based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
We modeled species activation as:
\begin{equation}
	v_{j} = \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
\end{equation}where $0\leq f_{ij}\left(x\right)\leq 1$ was a regulatory transfer function that calculated the influence of species $x_{i}$ on rate process $j$. 
Conversely, we modeled the inhibition (or repression) of rate process $j$ as:
\begin{equation}
	v_{j} = 1 - \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
\end{equation}Lastly, if rate process $j$ was subject to had both $m$ activating and $n$ inhibitory factors, we modeled the control term as:
\begin{equation}
	v_{j} = \min\left(u_{j},d_{j}\right)
\end{equation}where:
\begin{eqnarray}
	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right) \\
	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(x\right),\hdots,f_{nj}\left(x\right)\right)
\end{eqnarray}The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating, and inhibitory factors for process $j$. 
If a process had no control factors, we set $v_{j} = 1$.
There are many possible functional forms for $0\leq f_{ij}\left(x\right)\leq 1$. 
However, in this study, each individual transfer function took the form:
\begin{equation}\label{eqn:control-factor}
	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}x_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}x_{j}^{\eta}}
\end{equation}where $x_{j}$ denotes the abundance (or activity) of component $j$.  
The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
The model equations were encoded using the Octave programming language, and solved using the LSODE routine in Octave \citep{Octave:2014}.

\subsection*{Estimation of an ensemble of model parameters from experimental data.}
Model parameters were estimated by minimizing the squared difference between simulations and experimental data taken from ATRA-induced HL-60 cells:
\begin{equation}\label{eqn:objective-function}
	E_{j}(\mathbf{k}) = \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2 + \left(\frac{{\mathcal{M}^{\prime}_{ij}}-\max{y_{ij}}}{{\mathcal{M}^{\prime}_{ij}}}\right)^{2}
\end{equation}
The terms $\hat{\mathcal{M}}_{ij}$ and $\hat{y}_{ij}$ denote scaled experimental observations and simulation outputs from training set $j$. 
The quantity $i$ denoted the sampled time-index and $\mathcal{T}_{j}$ denoted the number of time points for experiment j. 
The first term in Eqn. \eqref{eqn:objective-function} quantified the relative error in the simulation. 
We used only immunoblot measurements for model training. 
Thus, we trained the model on the \emph{relative} change between bands within each training data set.
The read-out from the training immunoblots was band intensity where we assumed intensity was only loosely proportional to concentration.
Suppose we have the intensity for species $x$ at time $\{t_{1},t_{2},..,t_{n}\}$ in condition $j$. The scaled-value $\hat{\mathcal{M}}_{ij}$ would then be given by:
\begin{equation}\label{norm_exp_data}
\hat{\mathcal{M}}_{ij} = \frac{\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}{\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}
\end{equation}
Under this scaling $0\leq\hat{\mathcal{M}}_{ij}\leq{1}$ where $\hat{\mathcal{M}}_{ij}=0$ describes the lowest intensity band and $\hat{\mathcal{M}}_{ij}=1$ describes the highest intensity band. 
A similar scaling was defined for the simulation output. The second-term in the objective function ensured the proper concentration scale was estimated by the model.
In this study, we set the highest intensity band to $\mathcal{M}^{\prime}_{ij} = 10$ [AU] for all simulations.

We minimized the total model residual ($\sum{E_{j}}$) using Particle swarm optimization (PSO) \citep{PSO}.
PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
From each of these local points, a globally best error is identified. Both the local and global error 
are then used to update the parameter estimates of each particle using the rules:
\begin{eqnarray}
	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
$\mathcal{G}$ denotes the best solution found over the entire population of particles. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$, which was taken from XXX.

\section*{Acknowledgements}
The authors thank the anonymous reviewers for their helpful suggestions. 
We acknowledge the gracious financial support to J.V. by the National Science Foundation CAREER (CBET-0846876) for the support of R.T. and H.J.
In addition, we acknowledge support to A.Y. from the National Institutes of Health (CA 30555, CA152870) 
and a grant from New York State Stem Cell Science.  Lastly, we acknowledge the financial support to J.V. and A.Y. 
from the National Cancer Institute (\#U54CA143876). The content is solely the responsibility of the authors and does not necessarily 
represent the official views of the National Cancer Institute or the National Institutes of Health.
